---
title: "Biostat 203B Homework 4"
author: Juliet Yixuan Zhou
subtitle: Due Mar 12 @ 11:59PM
output:
  # ioslides_presentation: default
  html_document:
    toc: true
    toc_depth: 4
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```
                      
Display machine information:
```{r}
sessionInfo()
```
Load database libraries and the tidyverse frontend:
```{r}
library(tidyverse)
library(lubridate)
library(miceRanger)
library(readr)
```

## Q1. Missing data

Through the Shiny app developed in HW3, we observe abundant missing values in the MIMIC-IV ICU cohort we created. In this question, we use multiple imputation to obtain a data set without missing values.

0. Read following tutorials on the R package miceRanger for imputation: <https://github.com/farrellday/miceRanger>, <https://cran.r-project.org/web/packages/miceRanger/vignettes/miceAlgorithm.html>.

    A more thorough book treatment of the practical imputation strategies is the book [*_Flexible Imputation of Missing Data_*](https://stefvanbuuren.name/fimd/) by Stef van Buuren. 

1. Explain the jargon MCAR, MAR, and MNAR.

**Solution**: MCAR refers to data being Missing Completely At Random, which suggests that the causes of missing data are unrelated to the data itself. MAR refers to data being Missing At Random, which suggests that the probability of having missing data is the same within groups defined by the observed data, and might be different between groups. MNAR refers to data being Missing Not At Random, which suggests that the probability of having missing data varies for reasons that are unknown. 

2. Explain in a couple of sentences how the Multiple Imputation by Chained Equations (MICE) work.

**Solution**: For a dataset with p variables, replace the missing values for each variable with placeholders, e.g. mean of the not missing observations for each variable. Set the placeholders of one of these variables back to missing, and use this variable as the dependent variable in a regression model and all the other variables as predictors. Use this model to predict and replace missing values for this variable. Repeat this process for all the variables, and the repeat the whole process for several cycles. 

3. Perform a data quality check of the ICU stays data. Discard variables with substantial missingness, say >5000 `NA`s. Replace apparent data entry errors by `NA`s.

```{r eval=F}
icu <- read_rds("/Users/julietzhou/Documents/R/HW3_Shiny/icu_cohort.rds")

colSums(is.na(icu)) > 5000

for (i in colnames(icu)){
   print(class(df[[i]]))
}

```


4. Impute missing values by `miceRanger` (request $m=3$ datasets). This step is very computational intensive. Make sure to save the imputation results as a file.

5. Make imputation diagnostic plots and explain what they mean.

6. Obtain a complete data set by averaging the 3 imputed data sets.

## Q2. Predicting 30-day mortality

Develop at least two analytic approaches for predicting the 30-day mortality of patients admitted to ICU using demographic information (gender, age, marital status, ethnicity), first lab measurements during ICU stay, and first vital measurements during ICU stay. For example, you can use (1) logistic regression (`glm()` function), (2) logistic regression with lasso penalty (glmnet package), (3) random forest (randomForest package), or (4) neural network.

1. Partition data into 80% training set and 20% test set. Stratify partitioning according the 30-day mortality status.

2. Train the models using the training set.

3. Compare model prediction performance on the test set.
