---
title: "Biostat 203B Homework 1"
author: Juliet Yixuan Zhou
subtitle: Due Jan 22 @ 11:59PM
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Display machine information for reproducibility:
```{r, eval=T}
sessionInfo()
```

## Q1. Git/GitHub

**No handwritten homework reports are accepted for this course.** We work with Git and GitHub. Efficient and abundant use of Git, e.g., frequent and well-documented commits, is an important criterion for grading your homework.

1. Apply for the [Student Developer Pack](https://education.github.com/pack) at GitHub using your UCLA email.

**Solution**: Done.

2. Create a **private** repository `biostat-203b-2021-winter` and add `Hua-Zhou`, `Chris-German` and `ElvisCuiHan` as your collaborators with write permission.

**Solution**: Done.

3. Top directories of the repository should be `hw1`, `hw2`, ... Maintain two branches `master` and `develop`. The `develop` branch will be your main playground, the place where you develop solution (code) to homework problems and write up report. The `master` branch will be your presentation area. Submit your homework files (R markdown file `Rmd`, `html` file converted from R markdown, all code and data sets to reproduce results) in `master` branch.

**Solution**: Done.

4. After each homework due date, teaching assistant and instructor will check out your master branch for grading. Tag each of your homework submissions with tag names `hw1`, `hw2`, ... Tagging time will be used as your submission time. That means if you tag your `hw1` submission after deadline, penalty points will be deducted for late submission.

**Solution**: Done.

5. After this course, you can make this repository public and use it to demonstrate your skill sets on job market.


## Q2. Linux Shell Commands

1. This exercise (and later in this course) uses the [MIMIC-IV data](https://mimic-iv.mit.edu), a freely accessible critical care database developed by the MIT Lab for Computational Physiology. Follow the instructions at <https://mimic-iv.mit.edu/docs/access/> to (1) complete the CITI `Data or Specimens Only Research` course and (2) obtain the PhysioNet credential for using the MIMIC-IV data. Display the verification links to your completion report and completion certificate here. (Hint: The CITI training takes a couple hours and the PhysioNet credentialing takes a couple days; do not leave it to the last minute.)

**Solution**: Here is my [course completion report](citiprogram.org/verify/?ke004058f-7856-4ab9-bd6f-bedda1045e28-40365014) and my [course completion certificate](citiprogram.org/verify/?w367a1a4c-2e17-4855-a27c-b5cca9c074fc-40365014). The PhysioNet credential application has also been approved.

2. The `/usr/203b-data/mimic-iv/` folder on teaching server contains data sets from MIMIC-IV. Refer to <https://mimic-iv.mit.edu/docs/datasets/> for details of data files.  
    ```{bash}
    ls -l /usr/203b-data/mimic-iv
    ```
Please, do **not** put these data files into Git; they are big. Do **not** copy them into your directory. Do **not** decompress the gz data files. These create unnecessary big files on storage and are not big data friendly practices. Just read from the data folder `/usr/203b-data/mimic-iv` directly in following exercises. 

**Solution**: Noted. 

    Use Bash commands to answer following questions.

3. Display the contents in the folders `core`, `hosp`, `icu`. What are the functionalities of the bash commands `zcat`, `zless`, `zmore`, and `zgrep`? 

**Solution**: 

```{bash}
ls -l /usr/203b-data/mimic-iv/core
```
    
```{bash}
ls -l /usr/203b-data/mimic-iv/hosp
```
    
```{bash}
ls -l /usr/203b-data/mimic-iv/icu
```

The following information is obtained using the `man` command. 
zcat uncompresses a list of files writes the uncompressed data on standard output. 

zless and zmore are used to examine compressed or plain text files one screenful at a time on a soft-copy terminal. 

zgrep searches possibly compressed files for a regular expression. 

```{bash, eval=F}
man zcat
man zless
man zmore
man zgrep
```
    
4. What's the output of following bash script?

**Solution**:

The following code displays detailed information of files that end in `.gz` in the core folder. 
```{bash, eval=T}
for datafile in /usr/203b-data/mimic-iv/core/*.gz
  do
    ls -l $datafile
  done
```
Display the number of lines in each data file using a similar loop.

```{bash}
for datafile in /usr/203b-data/mimic-iv/core/*.gz
  do 
    wc -l $datafile
  done
```


4. Display the first few lines of `admissions.csv.gz`. How many rows are in this data file? How many unique patients (identified by `subject_id`) are in this data file? What are the possible values taken by each of the variable `admission_type`, `admission_location`, `insurance`, `language`, `marital_status`, and `ethnicity`? Also report the count for each unique value of these variables. (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `uniq`, `wc`, and so on.)

**Solution**:
Display the first five lines `admissions.csv.gz`. 
```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | head -5
```

Display the number of lines in this data file, which is 524521. 
```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | wc -l
```

Display the number of unique patients in this data file: print sorted `subject_id` from data file, then use uniq to find the number of unique occurrences.  

Note: the number generated from the following code, 257367, includes the header line as a unique entry too. 

```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, '{ print $1 }' | 
sort | uniq | wc -l
```
    
The code below fixes the header issue. Therefore, there should be 257366 unique patients in the admissions data file.  
```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, '{ print $1 }' | 
tail -n+2 | sort | uniq | wc -l
```

Display all possible values in the variable `admission_type` and their counts:
```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, '{ print $6 }' | 
tail -n+2 | sort | uniq -c
```

Display all possible values in the variable `admission_location` and their counts:
```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, '{ print $7 }' | 
tail -n+2 | sort | uniq -c
```

Display all possible values in the variable `insurance` and their counts:
```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, '{ print $9 }' | 
tail -n+2 | sort | uniq -c
```

Display all possible values in the variable `language` and their counts:
```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, '{ print $10 }' | 
tail -n+2 | sort | uniq -c
```

Display all possible values in the variable `marital_status` and their counts:
```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, '{ print $11 }' | 
tail -n+2 | sort | uniq -c
```

Display all possible values in the variable `ethnicity` and their counts:
```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk -F, '{ print $12 }' | 
tail -n+2 | sort | uniq -c
```

## Q3. Who's popular in Price and Prejudice

1. You and your friend just have finished reading *Pride and Prejudice* by Jane Austen. Among the four main characters in the book, Elizabeth, Jane, Lydia, and Darcy, your friend thinks that Darcy was the most mentioned. You, however, are certain it was Elizabeth. Obtain the full text of the novel from <http://www.gutenberg.org/cache/epub/42671/pg42671.txt> and save to your local folder. 
    ```{bash, eval=T}
    curl http://www.gutenberg.org/cache/epub/42671/pg42671.txt > pride_and_prejudice.txt
    ```
Do **not** put this text file `pride_and_prejudice.txt` in Git. Using a `for` loop, how would you tabulate the number of times each of the four characters is mentioned?

**Solution**:

Elizabeth is mentioned 634 times, Jane 294 times, Lydia 170 times, and Darcy 417 times.    
```{bash}
for name in Elizabeth Jane Lydia Darcy
  do
    grep -o $name pride_and_prejudice.txt | wc -l
  done
```
2. What's the difference between the following two commands?
    ```{bash eval=F}
    echo 'hello, world' > test1.txt
    ```
    and
    ```{bash eval=F}
    echo 'hello, world' >> test2.txt
    ```

**Solution**:
`>` directs the output of echo to the test1.txt file, while `>>` appends the output to the test2.txt file. 

3. Using your favorite text editor (e.g., `vi`), type the following and save the file as `middle.sh`:
    ```{bash eval=FALSE}
    #!/bin/sh
    # Select lines from the middle of a file.
    # Usage: bash middle.sh filename end_line num_lines
    head -n "$2" "$1" | tail -n "$3"
    ```
Using `chmod` make the file executable by the owner, and run 
    ```{bash eval=FALSE}
    ./middle.sh pride_and_prejudice.txt 20 5
    ```

Explain the output. Explain the meaning of `"$1"`, `"$2"`, and `"$3"` in this shell script. Why do we need the first line of the shell script?

**Solution**:

The middle.sh outputs the last few lines before a specified line number from the input file. Here, `"$1"` specifies the input file, e.g. `pride_and_prejudice.txt`. `"$2"` indicates the line number to end at, e.g. the 20th line, and `"$3"` indicates the number of last few lines to output, e.g. the last 5 lines. In the example, we want the last 5 lines ending at the 20th line of the pride_and_prejudice.txt file.


### Q4. More fun with Linux

Try these commands in Bash and interpret the results: `cal`, `cal 2021`, `cal 9 1752` (anything unusual?), `date`, `hostname`, `arch`, `uname -a`, `uptime`, `who am i`, `who`, `w`, `id`, `last | head`, `echo {con,pre}{sent,fer}{s,ed}`, `time sleep 5`, `history | tail`.

**Solution**: `cal` displays the calendar of the current month:
```{bash}
cal
```
`cal 2021` displays the calender of year 2021:
```{bash}
cal 2021
```
`cal 9 1752` displays the calendar for September of 1972. It is missing days 3 to 13, because these 11 days were skipped for leap year adjustments. 
```{bash}
cal 9 1752
```
`date` displays the current time and date.
```{bash}
date
```
`hostname` shows the system's host name. 
```{bash}
hostname
```

`arch` displays the machine hardware name. 
```{bash}
arch
```
`uname -a` prints system information. `-a` specifies printing all of the following information: kernel name, network and hostname, kernel release, kernel version, machine hardware name, processor type, hardware platform, and operating system. 
```{bash}
uname -a
```
`uptime` displays how long the system has been running. 
```{bash}
uptime
```
`who am i` displays the username. 
```{bash}
whoami
```
`who` displays the users who are logged on the server. 
```{bash}
who
```
`w` displays who is logged on and what they are doing on the server. 
```{bash}
w
```
`id` displays the user and group information for the current user. 
```{bash}
id
```

`last | head` displays the first ten lines of a list of last logged in users. 
```{bash}
last | head
```
`echo {con,pre}{sent,fer}{s,ed}` displays all 8 combinations from choosing one element from each of the three lists. 
```{bash}
echo {con,pre}{sent,fer}{s,ed}
```
`time` is used to time the execution of a script, and `sleep` is used to delay for a specified amount of time. Overall, `time sleep 5` asks to wait for 5 seconds before execution. 
```{bash}
time sleep 5
```
`history | tail`
```{bash}
echo "abc"
set -o history
history | tail
```

